\chapter{Statistical Commands}

\section{Zelig Commands}\label{s:commands}
  
\subsection{Quick Overview}\label{overview}
  
For any statistical model, Zelig does its work with a combination of
three commands.  

\begin{figure}[h!]
\caption{Main Zelig commands (solid arrows) and some options (dashed arrows)}
\label{3steps}
\begin{center}
\begin{displaymath}
\SelectTips{eu}{12} \UseTips 
\xymatrix@!C{\quad {\rm Imputation}
\ar@{-->}[dr] & & {\rm Matching} \ar@{-->}[dl] \\ 
 & *+[F-:<10pt>]{{\tt zelig()}} \ar@{->}[d] \ar@{-->}[r] &
{\tt summary()}\\ 
 & *+[F-:<10pt>]{{\tt setx()}}  \ar@{-->}[r] \ar@{->}[d] & {\tt whatif()}\\ 
& *+[F-:<10pt>]{{\tt sim()}} \ar@{-->}[dr] \ar@{-->}[dl] & \\
{\tt summary()} & &{\tt plot()} } 
\end{displaymath}
\end{center}
\end{figure}

\begin{enumerate}
\item Use \texttt{zelig()} to run the chosen statistical model on a
given data set, with a specific set of variables.  For standard
likelihood models, for example, this step estimates the coefficients,
other model parameters, and a variance-covariance matrix.  In
addition, you may choose from a variety of options:
\begin{itemize} 
\item Pre-processing:  Prior to calling {\tt zelig()}, you may
  choose from a variety of data pre-processing commands (matching or
  multiple imputation, for example) to make your statistical
  inferences more accurate. 
\item Summarize model: After calling {\tt zelig()}, you may summarize
the regression output using {\tt summary()}.  
\end{itemize}
\item Use \texttt{setx()} to set each of the explanatory variables to
  chosen (actual or counterfactual) values in preparation for
  calculating quantities of interest.  After calling {\tt setx()}, you
  may use \hlink{WhatIf}{/whatif/} to evaluate these choices by
  determining whether they involve interpolation (i.e., are inside the
  convex hull of the observed data) or extrapolation, as well as how
  far these counterfactuals are from the data.  Counterfactuals chosen
  in \texttt{setx()} that involve extrapolation far from the data can
  generate considerably more model dependence (see \cite{KinZen06a},
  \cite{KinZen06b}, \cite{StoKinZen05}).
\item Use \texttt{sim()} to draw simulations of your quantity of
  interest (such as a predicted value, predicted probability, risk
  ratio, or first difference) from the model.  (These simulations may
  be drawn using an asymptotic normal approximation (the default),
  bootstrapping, or other methods when available, such as directly
  from a Bayesian posterior.)  After calling {\tt sim()}, use any of
the following to summarize the simulations:
\begin{itemize}
\item The {\tt summary()} function gives a numerical display. For
multiple {\tt setx()} values, {\tt summary()} lets you summarize
simulations by choosing one or a subset of observations.
\item If the {\tt setx()} values consist of only one observation, {\tt
plot()} produces density plots for each quantity of interest.  
\end{itemize}
\end{enumerate}
Whenever possible, we use {\tt z.out} as the {\tt zelig()} output
object, \texttt{x.out} as the \texttt{setx()} output object, and {\tt
s.out} as the {\tt sim()} output object, but you may choose other
names.

\subsection{Examples}

\begin{itemize}
\item Use the \texttt{turnout} data set included with Zelig to
  estimate a logit model of an individual's probability of voting as
  function of race and age. Simulate the predicted probability of
  voting for a white individual, with age held at its mean:
\begin{verbatim}
> data(turnout)
> z.out <- zelig(vote ~ race + age, model = "logit", data = turnout)
> x.out <- setx(z.out, race = "white")
> s.out <- sim(z.out, x = x.out)
> summary(s.out)
\end{verbatim} 
\item Compute a first difference and risk ratio, changing education
  from 12 to 16 years, with other variables held at their means in the
  data:
\begin{verbatim}
> data(turnout)
> z.out <- zelig(vote ~ race + educate, model = "logit", data = turnout)
> x.low <- setx(z.out, educate = 12)
> x.high <- setx(z.out, educate = 16)
> s.out <- sim(z.out, x = x.low, x1 = x.high)
> summary(s.out)                                     # Numerical summary.
> plot(s.out)                                        # Graphical summary.
\end{verbatim}
\item Calculate expected values for every observation in your
  data set:
\begin{verbatim}
> data(turnout)
> z.out <- zelig(vote ~ race + educate, model = "logit", data = turnout)
> x.out <- setx(z.out, fn = NULL)
> s.out <- sim(z.out, x = x.out)
> summary(s.out)
\end{verbatim} 

\item Use five multiply imputed data sets from \cite{SchSla01} in an
  ordered logit model:
\begin{verbatim}
> data(immi1, immi2, immi3, immi4, immi5)
> z.out <- zelig(as.factor(ipip) ~ wage1992 + prtyid + ideol,
                 model = "ologit",
                 data = list(immi1, immi2, immi3, immi4, immi5))
\end{verbatim}

\item Use the nearest propensity score matching via {\it MatchIt}
  package, and then calculate the conditional average treatment effect
  of the job training program based on the linear regression model:
\begin{verbatim}
> library(MatchIt)
> data(lalonde)
> m.out <- matchit(treat ~ re74 + re75 + educ + black + hispan + age, 
                   data = lalonde, method = "nearest")
> m.data <- match.data(m.out)
> z.out <- zelig(re78 ~ treat + distance + re74 + re75 + educ + black +
                 hispan + age, data = m.data, model = "ls")
> x.out0 <- setx(z.out, fn = NULL, treat = 0)
> x.out1 <- setx(z.out, fn = NULL, treat = 1)
> s.out <- sim(z.out, x=x.out0, x1=x.out1)
> summary(s.out)
\end{verbatim}

\end{itemize}

\subsection{Details}

\begin{enumerate}
\item {\tt z.out <- zelig(formula, model, data, by = NULL, \dots)}
 
  The {\tt zelig()} command estimates a selected statistical model
  given the specified data.  You may name the output object
  (\texttt{z.out} above) anything you desire.  You must include three
  required arguments, in the following order:

  \begin{enumerate}

  \item \texttt{formula} takes the form {\tt y \~\, x1 + x2}, where
    {\tt y} is the dependent variable and {\tt x1} and {\tt x2} are
    the explanatory variables, and {\tt y}, {\tt x1}, and {\tt x2} are
    contained in the same dataset.  The {\tt +} symbol means
    ``inclusion'' not ``addition.''  You may include interaction terms
    in the form of {\tt x1*x2} without having to compute them in prior
    steps or include the main effects separately.  For example, R
    treats the formula {\tt y \~\, x1*x2} as {\tt y \~\, x1 + x2 +
      x1*x2}.  To prevent R from automatically including the separate
    main effect terms, use the {\tt I()} function, thus: {\tt y \~\,
      I(x1 * x2)}.

  \item \texttt{model} lets you choose which statistical model to run.
    You must put the name of the model in quotation marks, in the form
    {\tt model = "ls"}, for example.  See \Sref{s:models} for a list
    of currently supported models.

  \item \texttt{data} specifies the data frame containing the
    variables called in the formula, in the form {\tt data = mydata}.
    Alternatively, you may input multiply imputed datasets in the form
    {\tt data = list(data1, data2, \dots)}.\footnote{Multiple
      imputation is a method of dealing with missing values in your
      data which is more powerful than the usual list-wise deletion
      approach.  You can create multiply imputed datasets with a
      program such as
      \hlink{Amelia}{http://gking.harvard.edu/stats.shtml\#amelia};
      see \hlink{King, Honaker, Joseph, Scheve
        (2000)}{http://gking.harvard.edu/files/abs/evil-abs.shtml}.
    \label{mi}}  If you are working with matched data created using
  \hlink{MatchIt}{http://gking.harvard.edu/matchit/}, you may create a
  data frame within the {\tt zelig()} statement by using {\tt data =
    match.data(\dots)}.  In all cases, the data frame or MatchIt
  object must have been previously loaded into the working memory.
  
\item {\tt by} (an optional argument which is by default {\tt NULL})
  allows you to choose a factor variable (see \Sref{factors}) in the
  data frame as a subsetting variable.  For each of the unique
  strata defined in the {\tt by} variable, {\tt zelig()} does a
  separate run of the specified model.  The variable chosen should
  \emph{not} be in the formula, because there will be no variance in
  the {\tt by} variable in the subsets.  If you have one data set for
  all 191 countries in the UN, for example, you may use the {\tt by}
  option to run the same model 191 times, once on each country, all
  with a single {\tt zelig()} statement.  You may also use the {\tt
    by} option to run models on
  \hlink{MatchIt}{http://gking.harvard.edu/matchit/} subclasses.
  
\item The output object, {\tt z.out}, contains all of the options
  chosen, including the name of the data set.  Because data sets may
  be large, Zelig does not store the full data set, but only the name
  of the dataset.  Every time you use a Zelig function, it looks for
  the dataset with the appropriate name in working memory.  (Thus, it
  is critical that you do \emph{not} change the name of your data set,
  or perform any additional operations on your selected variables
  between calling {\tt zelig()} and {\tt setx()}, or between {\tt
    setx()} and {\tt sim()}.)
  
\item If you would like to view the regression output at this
  intermediate step, type {\tt summary(z.out)} to return the
  coefficients, standard errors, $t$-statistics and $p$-values.  We
  recommend instead that you calculate quantities of interest;
  creating {\tt z.out} is only the first of three steps in this task.
  \end{enumerate}
  
\item {\tt x.out <- setx(z.out, fn = list(numeric = mean, ordered =
    median, others = mode), data = NULL, cond = FALSE, \dots)}
 
  The {\tt setx()} command lets you choose values for the explanatory
  variables, with which {\tt sim()} will simulate quantities of
  interest.  There are two types of {\tt setx()} procedures:
\begin{itemize}
\item You may perform the usual \emph{unconditional} prediction (by
  default, {\tt cond = FALSE}), by explicitly choosing the values of
  each explanatory variable yourself or letting {\tt setx()} compute
  them, either from the data used to create {\tt z.out} or from a new
  data set specified in the optional {\tt data} argument.  You may
  also compute predictions for all observed values of your explanatory
  variables using {\tt fn = NULL}.
  
\item Alternatively, for advanced uses, you may perform
  \emph{conditional} prediction ({\tt cond = TRUE}), which predicts
  certain quantities of interest by conditioning on the observed value
  of the dependent variable.  In a simple linear regression model,
  this procedure is not particularly interesting, since the
  conditional prediction is merely the observed value of the dependent
  variable for that observation.  However, conditional prediction is
  extremely useful for other models and methods, including the
  following:
  \begin{itemize}
  \item In a matched sampling design, the sample average treatment
    effect for the treated can be estimated by computing the
    difference between the observed dependent variable for the treated
    group and their expected or predicted values of the dependent
    variable under no treatment \citep{HoImaKin05}.
  \item With censored data, conditional prediction will ensure that
    all predicted values are greater than the censored observed
    values \citep{KinAltBur90}.  
  \item In ecological inference models, conditional prediction
    guarantees that the predicted values are on the tomography line
    and thus restricted to the known bounds
    \citep{King97,AdoKinHer03}.
  \item The conditional prediction in many linear random effects (or
  Bayesian hierarchical) models is a weighted average of the
  unconditional prediction and the value of the dependent variable for
  that observation, with the weight being an estimable function of the
  accuracy of the unconditional prediction \citep[see][]{GelKin94a}.
  When the unconditional prediction is highly certain, the weight on
  the value of the dependent variable for this observation is very
  small, hence reducing inefficiency; when the unconditional
  prediction is highly uncertain, the relative weight on the
  unconditional prediction is very small, hence reducing bias.
  Although the simple weighted average expression no longer holds in
  nonlinear models, the general logic still holds and the mean square
  error of the measurement is typically reduced
  \citep[see][]{KinMurSal04}.
  \end{itemize}
  In these and other models, conditioning on the observed value of the
  dependent variable can vastly increase the accuracy of prediction
  and measurement.  
\end{itemize}
  
  The {\tt setx()} arguments for \textbf{unconditional} prediction are
  as follows:

   \begin{enumerate}

   \item {\tt z.out}, the {\tt zelig()} output object, must be
     included first.

   \item You can set particular explanatory variables to specified
   values.  For example:
\begin{verbatim}
> z.out <- zelig(vote ~ age + race, model = "logit", data = turnout)
> x.out <- setx(z.out, age = 30)
\end{verbatim}
   \texttt{setx()} sets the variables \emph{not} explicitly listed to
   their mean if numeric, and their median if ordered factors, and
   their mode if unordered factors, logical values, or character
   strings.  Alternatively, you may specify one explanatory variable
   as a range of values, creating one observation for every unique
   value in the range of values:\footnote{If you allow more than one variable to 
vary at a time, you risk confounding the predictive effect of the variables in 
question.} \begin{verbatim}
> x.out <- setx(z.out, age = 18:95)   
\end{verbatim}
   This creates 78 observations with with age set to 18 in the first
   observation, 19 in the second observation, up to 95 in the 78th
   observation.  The other variables are set to their default values,
   but this may be changed by setting {\tt fn}, as described next.
   
 \item Optionally, \texttt{fn} is a list which lets you to choose a different
   function to apply to explanatory variables of class
  \begin{itemize} 
    \item {\tt numeric}, which is {\tt mean} by default,
    \item {\tt ordered} factor, which is {\tt median} by default, and
    \item {\tt other} variables, which consist of logical variables,
      character string, and unordered factors, and are set to their
      {\tt mode} by default.
  \end{itemize}
  While any function may be applied to numeric variables, {\tt mean}
  will default to median for ordered factors, and mode is the only
  available option for other types of variables. In the special case,
  {\tt fn = NULL}, {\tt setx()} returns all of the observations.
   
 \item You cannot perform other math operations within the {\tt fn}
   argument, but can use the output from one call of {\tt setx} to
   create new values for the explanatory variables.  For example, to
   set the explanatory variables to one standard deviation below their
   mean:
\begin{verbatim}
> X.sd <- setx(z.out, fn = list(numeric = sd))     
> X.mean <- setx(z.out, fn = list(numeric = mean)) 
> x.out <- X.mean - X.sd           
\end{verbatim}
   
 \item Optionally, \texttt{data} identifies a new data frame (rather
   than the one used to create {\tt z.out}) from which the {\tt
     setx()} values are calculated.  You can use this argument to set
   values of the explanatory variables for hold-out or out-of-sample
   fit tests.
 \item The {\tt cond} is always {\tt FALSE} for unconditional
   prediction.
 \end{enumerate}
   
 If you wish to calculate risk ratios or first differences, call {\tt
   setx()} a second time to create an additional set of the values for
 the explanatory variables.  For example, continuing from the example
 above, you may create an alternative set of explanatory variables
 values one standard deviation above their mean:
\begin{verbatim}
> x.alt <- X.mean + X.sd
\end{verbatim}
 
 The required arguments for \textbf{conditional} prediction are as
 follows:
 \begin{enumerate}
 \item {\tt z.out}, the {\tt zelig()} output object, must be included
   first.
   
 \item {\tt fn}, which equals {\tt NULL} to indicate that all of the
   observations are selected.  You may only perform conditional
   inference on actual observations, not the mean of observations or
   any other function applied to the observations.  Thus, if {\tt fn}
   is missing, but {\tt cond = TRUE}, {\tt setx()} coerces {\tt fn =
     NULL}.

 \item {\tt data}, the data for conditional prediction.  
   
 \item {\tt cond}, which equals {\tt TRUE} for conditional prediction.
 \end{enumerate}
 Additional arguments, such as any of the variable names, are ignored
 in conditional prediction since the actual values of that observation
 are used.
 
\item {\tt s.out <- sim(z.out, x = x.out, x1 = NULL, num = c(1000,
    100), bootstrap = FALSE, bootfn = NULL, \dots)}
 
  The {\tt sim()} command simulates quantities of interest given the
  output objects from \texttt{zelig()} and \texttt{setx()}.  This
  procedure uses only the assumptions of the statistical model.  The
  {\tt sim()} command performs either unconditional or conditional
  prediction depending on the options chosen in {\tt setx()}.
  
  The arguments are as follows for \textbf{unconditional} prediction:
  \begin{enumerate}
  \item {\tt z.out}, the model output from {\tt zelig()}.
  \item {\tt x}, the output from the {\tt setx()} procedure performed
    on the model output.
  \item Optionally, you may calculate first differences by specifying
    {\tt x1}, an additional {\tt setx()} object.  For example, using
    the {\tt x.out} and {\tt x.alt}, you may generate first
    differences using:
\begin{verbatim}
> s.out <- sim(z.out, x = x.out, x1 = x.alt)
\end{verbatim}
    
  \item By default, the number of simulations, {\tt num}, equals 1000
    (or 100 simulations if bootstrap is selected), but this may be
    decreased to increase computational speed, or increased for
    additional precision.
    
  \item Zelig simulates parameters from classical \emph{maximum
      likelihood} models using asymptotic normal approximation to the
    log-likelihood.  This is the same assumption as used for
    frequentist hypothesis testing (which is of course equivalent to
    the asymptotic approximation of a Bayesian posterior with improper
    uniform priors).  See \hlink{King, Tomz, and Wittenberg
      (2000)}{http://gking.harvard.edu/files/abs/making-abs.shtml}.
    For \emph{Bayesian models}, Zelig simulates quantities of interest
    from the posterior density, whenever possible.  For \emph{robust
      Bayesian models}, simulations are drawn from the identified
    class of Bayesian posteriors.
  
  \item Alternatively, you may set {\tt bootstrap = TRUE} to simulate
    parameters using bootstrapped data sets.  If your dataset is
    large, bootstrap procedures will usually be more memory intensive
    and time-consuming than simulation using asymptotic normal
    approximation.  The type of bootstrapping (including the sampling
    method) is determined by the optional argument {\tt bootfn},
    described below.
  
  \item If {\tt bootstrap = TRUE} is selected, {\tt sim()} will
    bootstrap parameters using the default {\tt bootfn}, which
    re-samples from the data frame with replacement to create a
    sampled data frame of the same number of observations, and then
    re-runs {\tt zelig()} (inside {\tt sim()}) to create one set of
    bootstrapped parameters.  Alternatively, you may create a function
    outside the {\tt sim()} procedure to handle different bootstrap
    procedures.  Please consult {\tt help(boot)} for more
    details.\footnote{If you choose to create your own {\tt bootfn},
      it must include the the following three arguments: {\tt data},
      the original data frame; one of the sampling methods described
      in {\tt help(boot)}; and {\tt object}, the original {\tt
        zelig()} output object.  The alternative bootstrapping
      function must sample the data, fit the model, and extract the
      model-specific parameters.}
  \end{enumerate}
  
  For \textbf{conditional} prediction, {\tt sim()} takes only two
  required arguments:
  \begin{enumerate}
  \item {\tt z.out}, the model output from {\tt zelig()}.
  \item {\tt x}, the conditional output from {\tt setx()}.  
  \item Optionally, for duration models, {\tt cond.data}, which is the
    {\tt data} argument from {\tt setx()}.  For models for duration
    dependent variables (see \Sref{duration}), {\tt sim()} must impute
    the uncensored dependent variables before calculating the average
    treatment effect.  Inputting the {\tt cond.data} allows {\tt
      sim()} to generate appropriate values.
  \end{enumerate}
  Additional arguments are ignored or generate error messages.  

\end{enumerate}

\subsubsection*{Presenting Results}
\begin{enumerate}
\item Use {\tt summary(s.out)} to print a summary of your simulated
  quantities.  You may specify the number of significant digits as:
\begin{verbatim}
> print(summary(s.out), digits = 2)   
\end{verbatim}
\item Alternatively, you can plot your results using
  \texttt{plot(s.out)}.
\item You can also use \texttt{names(s.out)} to see the names and a
  description of the elements in this object and the {\tt \$} operator
  to extract particular results.  For most models, these are: {\tt
    s.out\$qi\$pr} (for predicted values), {\tt s.out\$qi\$ev} (for
  expected values), and {\tt s.out\$qi\$fd} (for first differences in
  expected values).  For the logit, probit, multinomial logit, ordinal
  logit, and ordinal probit models, quantities of interest also
  include {\tt s.out\$qi\$rr} (the risk ratio).
\end{enumerate} 

\section{Supported Models}\label{s:models}

We list here all models impelmented in Zelig, organized by the nature
of the dependent variable(s) to be predicted, explained, or described.

\begin{enumerate}
\item {\bf Continuous Unbounded} dependent variables can take any
real value in the range $(-\infty, \infty)$. While most of these
models take a continuous dependent variable, Bayesian factor analysis takes multiple
continuous dependent variables.  

   \begin{enumerate}
   \item {\tt "ls"}: The {\it linear least-squares} (see \Sref{ls})
     calculates the coefficients that minimize the sum of squared
     residuals.  This is the usual method of computing linear
     regression coefficients, and returns unbiased estimates of
     $\beta$ and $\sigma^2$ (conditional on the specified model).
   \item {\tt "normal"}: The {\it Normal} (see \Sref{normal}) model
     computes the maximum-likelihood estimator for a Normal
     stochastic component and linear systematic component.  The
     coefficients are identical to \texttt{ls}, but the
     maximum likelihood estimator for $\sigma^2$ is consistent
     but biased.
   \item {\tt "normal.bayes"}:  The {\it Bayesian Normal} regression
model (\Sref{normal.bayes}) is similar to maximum likelihood Gaussian
regression, but makes valid small sample inferences via draws from the
exact posterior and also allows for priors.  


  \item {\tt "tobit"}:  The {\it tobit} regression model (see 
  \Sref{tobit}) is a Normal distribution with left-censored observations.

  \item {\tt "tobit.bayes"}: The {\it Bayesian tobit} distribution
  (see \Sref{tobit.bayes}) is a Normal distribution that has either
  left and/or right censored observations.  

   \item {\tt "factor.bayes"}: The {\it Bayesian factor analysis} model (see
\Sref{factor.bayes}) estimates multiple observed continuous dependent variables
as a function of latent explanatory variables.  

   \end{enumerate}
 \item {\bf Dichotomous} dependent variables consist of two discrete
values, usually $(0,1)$.  
   \begin{enumerate}
   \item {\tt "logit"}: {\it Logistic regression} (see \Sref{logit})
     specifies $\Pr(Y=1)$ to be a(n inverse) logistic transformation
     of a linear function of a set of explanatory variables.
   \item {\tt "relogit"}: The {\it rare events logistic} regression
     option (see \Sref{relogit}) estimates the same model as the
     logit, but corrects for bias due to rare events (when one of the
     outcomes is much more prevalent than the other).  It also
     optionally uses prior correction to correct for choice-based
     (case-control) sampling designs.
   \item {\tt "logit.bayes"}: {\it Bayesian logistic regression} (see
\Sref{logit.bayes}) is similar to maximum likelihood logistic
regression, but makes valid small sample inferences via draws from the
exact posterior and also allows for priors.  
   \item {\tt "probit"}: {\it Probit regression} (see \Sref{probit})
     Specifies $\Pr(Y=1)$ to be a(n inverse) CDF normal transformation
     as a linear function of a set of explanatory variables.
   \item {\tt "probit.bayes"}: {\it Bayesian probit} regression (see
\Sref{probit.bayes}) is similar to maximum likelihood probit
regression, but makes valid small sample inferences via draws from the
exact posterior and also allows for priors. 
   \item {\tt "blogit"}: The {\it bivariate logistic} model (see
     \Sref{blogit}) models $\Pr(Y_{i1} = y_1, Y_{i2} = y_2)$ for
$(y_1, y_2) = {(0,0), (0,1), (1,0), (1,1)}$ according 
to a bivariate logistic density.
   \item {\tt "bprobit"}: The {\it bivariate probit} model (see
     \Sref{bprobit}) models $\Pr(Y_{i1} = y_1, Y_{i2} = y_2)$ for
$(y_1, y_2) = {(0,0), (0,1), (1,0), (1,1)}$ according to a bivariate normal density.  
   \item {\tt "irt1d"}: The {\it one-dimensional item response} model
(see \Sref{irt1d}) takes multiple dichotomous dependent variables and models
them as a function of \emph{one} latent (unobserved) explanatory variable.  
   \item {\tt "irtkd"}:  The {\it k-dimensional item response} model
(see \Sref{irtkd}) takes multiple dichotomous dependent variables and models
them as a function of $k$ latent (unobserved) explanatory variables.  
\end{enumerate} 
 \item {\bf  Ordinal} are used to model
   ordered, discrete dependent variables.  The values of the outcome
   variables (such as kill, punch, tap, bump) are ordered, but the
   distance between any two successive categories is not known
   exactly.  Each dependent variable may be thought of as linear, with
one continuous, unobserved dependent variable observed through a mechanism
   that only returns the ordinal choice.
   \begin{enumerate}
   \item {\tt "ologit"}: The {\it ordinal logistic} model (see
     \Sref{ologit}) specifies the stochastic component of the
     unobserved variable to be a standard logistic distribution.
   \item {\tt "oprobit"}: The {\it ordinal probit} distribution (see
     \Sref{oprobit}) specifies the stochastic component of the
     unobserved variable to be standardized normal.
      \item {\tt "oprobit.bayes"}: {\it Bayesian ordinal probit} model
(see \Sref{oprobit.bayes}) is similar to ordinal probit
regression, but makes valid small sample inferences via draws from the
exact posterior and also allows for priors.  
     \item {\tt "factor.ord"}: {\it Bayesian ordered factor analysis}
(see \Sref{factor.ord}) models observed, ordinal dependent variables
as a function of latent explanatory variables.  
   \end{enumerate}
 \item {\bf Multinomial} dependent variables are unordered, discrete
   categorical responses.  For example, you could model an
   individual's choice among brands of orange juice or among
   candidates in an election.
    \begin{enumerate}
    \item {\tt "mlogit"}: The {\it multinomial logistic} model (see
      \Sref{mlogit}) specifies categorical responses distributed
      according to the multinomial stochastic component and logistic
      systematic component.
    \item {\tt "mlogit.bayes"}: {\it Bayesian multinomial logistic} regression (see
\Sref{mlogit.bayes}) is similar to maximum likelihood multinomial logistic
regression, but makes valid small sample inferences via draws from the
exact posterior and also allows for priors. 
% \item {\tt "mprobit"}: The {\it multinomial probit} model (see
%      \Sref{mprobit}) specifies categorical responses distributed
%      according to the multinomial stochastic component and Normal
%      systematic component.
    \end{enumerate}
  \item {\bf Count} dependent variables are non-negative integer
values, such as the number of presidential vetoes or the number of
photons that hit a detector. 
    \begin{enumerate}
    \item {\tt "poisson"}: The {\it Poisson} model (see
      \Sref{poisson}) specifies the expected number of events that
      occur in a given observation period to be an exponential
      function of the explanatory variables.  The Poisson stochastic
      component has the property that, $\lambda = \textrm{E}(Y_i|X_i)
      = \textrm{V}(Y_i|X_i)$.
    \item {\tt "poisson.bayes"}: {\it Bayesian Poisson} regression (see
\Sref{poisson.bayes}) is similar to maximum likelihood Poisson
regression, but makes valid small sample inferences via draws from the
exact posterior and also allows for priors. 
    \item {\tt "negbin"}: The {\it negative binomial} model (see
      \Sref{negbin}) has the same systematic component as the Poisson,
      but allows event counts to be over-dispersed, such that
      $\textrm{V}(Y_i|X_i) > \textrm{E}(Y_i|X_i)$.
    \end{enumerate}

  \item {\bf Continuous Bounded}\label{duration} dependent variables
that are continuous only over a certain range, usually $(0, \infty)$.
In addition, some models (exponential, lognormal, and
Weibull) are also censored for values greater than some censoring
point, such that the dependent variable has some units fully observed
and others that are only partially observed (censored).  

 \begin{enumerate}

   \item {\tt "gamma"}: The {\it Gamma} model (see \Sref{gamma}) for
     positively-valued, continuous dependent variables that are fully
observed (no censoring).  

   \item {\tt "exp"}: The {\it exponential} model (see \Sref{exp}) for
   right-censored dependent variables assumes that the hazard function
   is constant over time.  For some variables, this may be an unrealistic
   assumption as subjects are more or less likely to fail the longer
   they have been exposed to the explanatory variables.

  \item {\tt "weibull"}: The {\it Weibull} model (see \Sref{weibull})
  for right-censored dependent variables relaxes the
  assumption of constant hazard by including an additional scale
  parameter $\alpha$: If $\alpha > 1$, the risk of failure increases
  the longer the subject has survived; if $\alpha < 1$, the risk of
  failure decreases the longer the subject has survived.  While {\tt
  zelig()} estimates $\alpha$ by default, you may optionally fix
  $\alpha$ at any value greater than 0.  Fixing $\alpha = 1$ results
  in an exponential model.

  \item {\tt "lognorm"}: The {\it log-normal} model (see
  \Sref{lognorm}) for right-censored duration dependent variables
  specifies the hazard function non-monotonically, with increasing
  hazard over part of the observation period and decreasing hazard
  over another.

\end{enumerate}

\item {\bf Mixed} dependent variables include models that take more
than one dependent variable, where the dependent variables come from
two or more of categories above.  (They do not need
to be of a homogeneous type.)
\begin{enumerate}
\item The {\it Bayesian mixed factor analysis} model, in contrast to the Bayesian factor
analysis model and ordinal factor analysis model, can model both types
of dependent variables as a function of latent explanatory variables.
\end{enumerate}

%  \item {\bf Models for Contingency Tables}
%     \begin{enumerate}
%     \item {\tt "mloglm"}: Use a {\it multinomial log-linear} model (see
%       \Sref{mloglm}) to estimate log-linear models for contingency
%       tables.  These models use raw counts in tables as the dependent
%       variable and apply a Poisson regression model.  Many of the rich
%       set of models that result are asymptotically equavlent to
%       multinomial logit models.  
%     \end{enumerate}
\end{enumerate}

\section{Replication Procedures} 

A large part of any statistical analysis is documenting your work such
that given the same data, anyone may replicate your results.  In
addition, many journals require the creation and dissemination of
``replication data sets'' in order that others may replicate your
results (see \hlink{King,
1995}{http://gking.harvard.edu/data.shtml\#repl}\nocite{King95}).
Whether you wish to create replication materials for your own records,
or contribute data to others as a companion to your published work,
Zelig makes this process easy.

\subsection{Saving Replication Materials}

Let {\tt mydata} be your final data set, {\tt z.out} be your {\tt
  zelig()} output, and {\tt s.out} your {\tt sim()} output.  To save
all of this in one file, type:
\begin{verbatim}
> save(mydata, z.out, s.out, file = "replication.RData")
\end{verbatim}
This creates the file replication.RData in your working directory.
You may compress this file using zip or gzip tools.

If you have run several specifications, all of these estimates may be
saved in one .RData file.  Even if you only created quantities of
interest from one of these models, you may still save all the
specifications in one file.  For example:
\begin{verbatim}
> save(mydata, z.out1, z.out2, s.out, file = "replication.RData")
\end{verbatim}

Although the .RData format can contain data sets as well as output
objects, it is not the most space-efficient way of saving large data
sets.  In an uncompressed format, ASCII text files take up less space
than data in .RData format.  (When compressed, text-formatted data is
still smaller than .RData-formatted data.)  Thus, if you have more
than 100,000 observations, you may wish to save the data set
separately from the Zelig output objects.  To do this, use the {\tt
  write.table()} command.  For example, if {\tt mydata} is a data
frame in your workspace, use {\tt write.table(mydata, file =
  "mydata.tab")} to save this as a tab-delimited ASCII text file.  You
may specify other delimiters as well; see {\tt
  help.zelig("write.table")} for options.

\subsection{Replicating Analyses}

If the data set and analyses are all saved in one .RData file, located
in your working directory, you may simply type:
\begin{verbatim}
> load("replication.RData")                   # Loads the replication file.  
> z.rep <- repl(z.out)                        # To replicate the model only. 
> s.rep <- repl(s.out)                        # To replicate the model and 
                                              #  quantities of interest.  
\end{verbatim}
By default, {\tt repl()} uses the same options used to create the
original output object.  Thus, if the original {\tt s.out} object used
bootstrapping with 245 simulations, the {\tt s.rep} object will
similarly have 245 bootstrapped simulations.  In addition, you may use
the {\tt prev} option when replicating quantities of interest to reuse
rather than recreate simulated parameters.  Type {\tt
  help.zelig("repl")} to view the complete list of options for
{\tt repl()}.

If the data were saved in a text file, use {\tt read.table()} to load
the data, and then replicate the analysis:
\begin{verbatim}
> dat <- read.table("mydata.tab", header = TRUE)  # Where `dat' is the same
> load("replication.RData")                       #   as the name used in 
> z.rep <- repl(z.out)                            #   `z.out'.
> s.rep <- repl(s.out)  
\end{verbatim}
If you have problems loading the data, please refer to \Sref{load.data}.

Finally, you may use the {\tt identical()} command to ensure that the
replicated regression output is in every way identical to the original
{\tt zelig()} output.\footnote{The {\tt identical()} command checks
  that numeric values are identical to the maximum number of decimal
  places (usually 16), and also checks that the the two objects have
  the same class (numeric, character, integer, logical, or factor).
  Refer to {\tt help(identical)} for more information.}  For example:
\begin{verbatim}
> identical(z.out$coef, z.rep$coef)              # Checks the coefficients.
\end{verbatim}
Simulated quantities of interest will vary from the original
quantities if parameters are re-simulated or re-sampled.  If you wish to
use {\tt identical()} to verify that the quantities of interest are
identical, you may use
\begin{verbatim}
# Re-use the parameters simulated (and stored) in the original sim() output.
> s.rep <- repl(s.out, prev = s.out$par) 

# Check that the expected values are identical.  You may do this for each qi.
> identical(s.out$qi$ev, s.rep$qi$ev) 
\end{verbatim} %$


